High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-07-23 doc/asplos2013/impl.tex
minor edits to section 3

There are currently two implementations of our interface:
one that works only on shared memory machines and another that
runs on larges clusters with both CPUs and GPUs.  The shared-memory
implementation uses the POSIX threads library\cite{PTHREADS} (Pthreads)
and is primarily used for debugging.
The implementation for heterogeneous clusters also uses Pthreads as 
well as the CUDA runtime library for GPUs\cite{CUDA} and the GASNet
API for large clusters\cite{GASNET07}.

The heterogeneous implementation
models a cluster with CPUs and GPUs as having two types of processors 
and four types of memory.  Every CPU core is presented as a different {\tt CPU} processor
and every GPU is a {\tt GPU} processor.  This matches the scheduling granularity available
in the Pthreads and CUDA APIs respectively.  The first type of memory is the 
system memory that is accessible by every CPU core on a given node.  The second is the 
framebuffer memory on each GPU (and accessible only by that GPU).  The third type
of memory is system memory on a node that has been made accessible to the GPU(s) on
that node as well as the CPU cores, commonly referred to as {\em zero-copy memory}.
The final type of memory is the portion of system memory on each node that has been registered with
the GASNet runtime to allow {\em remote memory access} (RMA) by other nodes in the cluster.
We refer to this memory as {\em GASNet memory}.
%% For every node in the system
%% there is a memory corresponding to the DRAM associated with that node.  For
%% every GPU in the system there are two memories corresponding to the framebuffer
%% memory and the zero-copy memory associated with that GPU \footnote{Zero-copy 
%% memory contains pages that are mapped in both framebuffer and node memory with
%% coherence maintained by the PCI-E protocol.}. The last memory is 
%% a global GASNet memory that represents pages that have been registered 
%% with all nodes in the system for supporting remote memory accesses (RMAs).
%% The GASNet memory provides the illusion of global memory.  CPU processors
%% can directly access their node memory, all zero-copy memories for GPUs on
%% their node, and the GASNet memory.  GPU processors can only access their
%% framebuffer memory and their zero-copy memory.  Copies are permitted
%% between all pairs of memory except between any framebuffer memory and
%% the GASNet memory.

In addition to the RMA capabilities, GASNet provides {\em active messages} for 
inter-node communication.  Active messages consist of a command and payload that
are sent by one node to another without any previous coordination.  Upon arrival
at a destination node, a handler routine is invoked to process an active message.
%The handler may issue a response to the sender, but this is optional, and avoided
%as much as possible in our implementation due to the latencies involved in waiting
%for a response.  
One example of a case where we use active messages involves the {\tt spawn} call on a {\tt Processor}
object.  If the processor is not on the same node as the requestor, the information
%about the task (e.g. the task ID and arguments, the actual target {\tt Processor},
%the prerequisite event (if any), and the completion event ID) 
for the task is sent in an
active message to the node containing the target processor.  The recipient node
unpacks the message, and either places the task on the ready queue for the target
processor or marks it as being deferred until the prerequisite event has triggered.
%No response is sent to the original requestor.  The requestor already knows the event
%that corresponds to the completion of this task (since it was provided in the active
%message) - there is no utility in knowing exactly when the active message itself has
%been processed.

@@

%Due to space constraints, we will not cover every part of our implementation in
%detail, instead limiting our discussion to the three most interesting aspects:
%events, locks, and reduction instances.
Due to space constraints we cover only a small part of the implementation in detail.
We limit our discussion to three interesting aspects:
events, locks, and reduction instances.

@@

%At event creation time, the owning node allocates a data structure to track the state of the event
%(i.e. triggered or not) as well as record a list of the operations that are known to be dependent
%on the event (e.g. initiation of a copy operation, placing a task on a processor's ready queue, waking
%up a waiting task).  However, the owning node only keeps a list of the dependent operations from the same
%node.  Every other node also allocates a corresponding data structure (the first time the event
%is referenced) to remember if the event has already triggered and to record its local dependent operations.
%Arbitrarily many dependent operations from a
%single node are aggregated into a single {\em event subscription} active message that is sent to the
%owning node.  The owning node keeps a bitmask of which other nodes have subscribed to the event, and when the
%event finally does trigger, a single {\em event trigger} active message is sent to each subscribing node,
%which notes that the event has triggered (in case queries come after the trigger) and executes the list
%of local operations that were dependent on that event.  (In the case that the triggering of the event has
%occurred before the reception of an event subscription, a trigger message is sent immediatel to the 
%new subscriber.)
When a new event is created, the owning node allocates a data structure to track the state of
the event ({\em triggered} or {\em untriggered}) as well as to record a list of operations dependent on the
event from the same node called {\em waiters} (e.g. copy operation or task launch).  The first operation contingent on an event
from a remote node will create the same data structure on the remote node.  A single
{\em event subscription} active message is then sent from the remote node to the owning node indicating
that the remote node should be informed when the event triggers.  Arbitrarily many additional operations on
the remote node can then be aggregated as waiters onto the remote data structure without any additional communication.
When the event does trigger, the owner node notifies all local waiters and
sends a single {\em event trigger} active message to each node from which it has seen an event subscription
to notify remote waiters.
In the case where an event triggers while a subscription message is in flight from a remote node, the owner node will
immediately respond with a trigger message.

@@

The data structures used to track an event cannot be freed until all operations on that event have been
performed.  Creation and triggering can each happen only once, but there can be an arbitrary number of operations
that are dependent on an event.   Furthermore, some operations may not be requested until long after the
event has triggered.  Other systems incorporating events address this by reference counting event
handles\cite{Khronos:OpenCL}, but such reference counting adds both client and runtime overhead even when
limited to a single node; further overhead and complexity would be expected for a cluster-level
reference-counting approach.

%Instead of focusing on freeing event data structures, our implementation aggressively recycles them, 
%needing fewer total event data structures than a referencing counting implementation while also eliminating
%the overhead of reference counting.  The key observation is that one {\em generational event} data structure
% can efficiently capture the state of one {\em untriggered} event and a very large number (e.g. $2^{32}-1$)
%of already-triggered events.  An event's handle is expanded to include its generation number as well as the
%identifier for the underlying generational event.  In addition to the triggered-or-not state and list of
%dependent operations
%for the most recent generation, the owning node also remembers how many previous generations have already
%been triggered.  (There is no need to keep a list of dependent operations on previous generations.  Any
%new operation that is dependent on a generation that is known to be already triggered can immediately be
%executed.)  When an event is created, any generational event for whom the most recent generation has 
%triggered can be reused - the generation is increased by one and the state is returned to ``untriggered.''
%As before, this can be done with no inter-node communication.
Instead of attempting to free event data structures, our implementation aggressively recycles them.
Compared to a reference counting scheme our implementation requires fewer total event data structures
and has no client/runtime overhead.  The key observations is that one {\em generational event}
data structure can efficiently capture the state of one untriggered event and a very large 
number (e.g. $2^{32}-1$) of already-triggered events.  To accomplish this, we extend each event
handle to include its generation number as well as the identifier for its generational event.  Each
generational event remembers how many previous generations have already been triggered.  Any
new operation that is dependent on a generation that is known to already be triggered can immediately
be executed.  A generational event can be reused for a new generation as soon as the current generation has triggered. To
create a new event, a node finds a generational event in the triggered state,
increases the generation by one, and sets the generational event's state to untriggered.  As before, 
this can be done with no inter-node communication.

@@

Like events, deferred locks, or locks from here on, are created on demand by the node on which the creation request was made.  The lock
handles are also statically divided across the nodes, allowing the creation to be done without inter-node
communication.  Unlike events whose ownership is static, locks may migrate; the {\em creating}
node is the initial owner, but that ownership can be transferred to other nodes.

@@

When a lock request is made, an event is created to track when the grant occurs.  The current node then
examines its copy of the lock data structure to determine if it is the owner.
%(creating it if this the first operation for that lock on that node).  
If the current node is the owner and the lock isn't
held, the lock is granted immediately and the event is triggered.  If the lock is held, the event is added
to the list of local waiters.  Note that the event associated with
the lock request is the only data that must be stored.
%(This is the only thing that is stored.  The actual ``requestor'' of the lock
%is implicitly captured in which operations are dependent on the lock grant event.)  
If the current node isn't the owner, a {\em lock request} active message is
sent to the most recently known owner.  If the receiving node is no
longer the owner, it forwards the message on to the node it has recorded as the owner.
%(If that information proves to be stale, the request is forwarded
%by the receiving node, eventually catching up to the current owner.)  
If the current owner's status shows
the lock is currently held, the bit for the requesting node is set in the remote waiters bitmask.  If the lock is
not held, the ownership of the lock is given to the requesting node via a {\em lock transfer} active
message.  A lock transfer message includes the bitmask of remote waiters and an up-to-date copy 
of the lock's payload.  The inclusion of the payload in the active message is the reason for the 
4KB size limit specified on payloads in Section~\ref{subsec:locks}.

@@

The first kind of reduction-only physical region is a {\em reduction fold instance}, which may only be
used for reduction operations that support a {\em fold} operation (see Listing~\ref{lst:regionapi}).  
A reduction fold instance is similar
to a normal physical region in that it is implemented as an array, indexed by the same element indices used
for the normal physical regions.  The difference is that each array element is a value of the reduction's
{\tt RHS} type which is generally the same size or smaller than the actual element type of the region 
(i.e. the {\tt LHS} type).  When the client requests a reduction operation to a particular location, 
the supplied right-hand-side value is folded in
to the corresponding location in the array.  An arbitrary number of reductions can be done to each location.
When the reduction fold instance is copied back to a normal instance, the contents of the reduction instance
are transferred in a large block to the location of the normal instance.  On the receiving side
the runtime automatically applies the reduction instance to the destination region
by invoking the reduction function once for each location in the region via a cache-friendly 
linear sweep over the memory.

>>>>

There are currently two implementations of our interface:
one that works only on shared memory machines and another that
runs on large clusters with both CPUs and GPUs.  The shared-memory
implementation uses the POSIX threads library\cite{PTHREADS} (Pthreads)
and is primarily used for debugging.
The implementation for heterogeneous clusters uses Pthreads as 
well as the CUDA runtime library for GPUs\cite{CUDA} and the GASNet
API for large clusters\cite{GASNET07}.

The heterogeneous implementation
models a cluster with CPUs and GPUs as having two kinds of processors 
and four kinds of memory.  Every CPU core is presented as a different {\tt CPU} processor
and every GPU is a {\tt GPU} processor.  This matches the scheduling granularity available
in the Pthreads and CUDA APIs respectively.  The first kind of memory is the 
system memory that is accessible by every CPU core on a given node.  The second is the 
framebuffer memory on each GPU (and accessible only by that GPU).  The third kind
of memory is system memory on a node that has been made accessible to the GPU(s) on
that node as well as the CPU cores, commonly referred to as {\em zero-copy memory}.
The final kind of memory is the portion of system memory on each node that has been registered with
the GASNet runtime to allow {\em remote memory access} (RMA) by other nodes in the cluster.
We refer to this memory as {\em GASNet memory}.
%% For every node in the system
%% there is a memory corresponding to the DRAM associated with that node.  For
%% every GPU in the system there are two memories corresponding to the framebuffer
%% memory and the zero-copy memory associated with that GPU \footnote{Zero-copy 
%% memory contains pages that are mapped in both framebuffer and node memory with
%% coherence maintained by the PCI-E protocol.}. The last memory is 
%% a global GASNet memory that represents pages that have been registered 
%% with all nodes in the system for supporting remote memory accesses (RMAs).
%% The GASNet memory provides the illusion of global memory.  CPU processors
%% can directly access their node memory, all zero-copy memories for GPUs on
%% their node, and the GASNet memory.  GPU processors can only access their
%% framebuffer memory and their zero-copy memory.  Copies are permitted
%% between all pairs of memory except between any framebuffer memory and
%% the GASNet memory.

In addition to the RMA capabilities, GASNet provides {\em active messages} for 
inter-node communication.  Active messages consist of a command and payload that
are sent by one node to another without any previous coordination.  Upon arrival
at a destination node, a handler routine is invoked to process the message.
%The handler may issue a response to the sender, but this is optional, and avoided
%as much as possible in our implementation due to the latencies involved in waiting
%for a response.  
One example of a case where we use active messages involves the {\tt spawn} call on a {\tt Processor}
object.  If the processor is not on the same node as the requestor, the information
%about the task (e.g. the task ID and arguments, the actual target {\tt Processor},
%the prerequisite event (if any), and the completion event ID) 
for the task is sent in an
active message to the node containing the target processor.  The recipient node
unpacks the message, and either places the task on the ready queue for the target
processor or marks it as being deferred until the prerequisite event has triggered.
%No response is sent to the original requestor.  The requestor already knows the event
%that corresponds to the completion of this task (since it was provided in the active
%message) - there is no utility in knowing exactly when the active message itself has
%been processed.

@@

%Due to space constraints, we will not cover every part of our implementation in
%detail, instead limiting our discussion to the three most interesting aspects:
%events, locks, and reduction instances.
Due to space constraints we cover only the most interesting parts of the implementation in detail.
We limit our discussion to three aspects:
events, locks, and reduction instances.

@@

%At event creation time, the owning node allocates a data structure to track the state of the event
%(i.e. triggered or not) as well as record a list of the operations that are known to be dependent
%on the event (e.g. initiation of a copy operation, placing a task on a processor's ready queue, waking
%up a waiting task).  However, the owning node only keeps a list of the dependent operations from the same
%node.  Every other node also allocates a corresponding data structure (the first time the event
%is referenced) to remember if the event has already triggered and to record its local dependent operations.
%Arbitrarily many dependent operations from a
%single node are aggregated into a single {\em event subscription} active message that is sent to the
%owning node.  The owning node keeps a bitmask of which other nodes have subscribed to the event, and when the
%event finally does trigger, a single {\em event trigger} active message is sent to each subscribing node,
%which notes that the event has triggered (in case queries come after the trigger) and executes the list
%of local operations that were dependent on that event.  (In the case that the triggering of the event has
%occurred before the reception of an event subscription, a trigger message is sent immediatel to the 
%new subscriber.)
When a new event is created, the owning node allocates a data structure to track the state of
the event ({\em triggered} or {\em untriggered}) as well as to record a list of dependent operations
(e.g. copy operation or task launch) from the same node (called {\em waiters}).
The first operation dependent on an event
from a remote node will allocate the same data structure on the remote node.  A single
{\em event subscription} active message is then sent from the remote node to the owning node indicating
that the remote node should be informed when the event triggers.  Arbitrarily many additional operations on
the remote node can then be added to the list of waiters without any additional communication.
When the event does trigger, the owner node notifies all local waiters and
sends a single {\em event trigger} active message to each node from which it has seen an event subscription
to notify remote waiters.
In the case where an event triggers while a subscription message is in flight from a remote node, the owner node will
immediately respond with a trigger message.

@@

The data structure used to track an event cannot be freed until all operations on that event have been
performed.  Creation and triggering can each happen only once, but there can be an arbitrary number of operations
that are dependent on an event.   Furthermore, some operations may not be requested until long after the
event has triggered.  Other systems incorporating events address this by reference counting event
handles\cite{Khronos:OpenCL}, but such reference counting adds both client and runtime overhead even when
limited to a single node; further overhead and complexity would be expected for a cluster-level
reference-counting approach.

%Instead of focusing on freeing event data structures, our implementation aggressively recycles them, 
%needing fewer total event data structures than a referencing counting implementation while also eliminating
%the overhead of reference counting.  The key observation is that one {\em generational event} data structure
% can efficiently capture the state of one {\em untriggered} event and a very large number (e.g. $2^{32}-1$)
%of already-triggered events.  An event's handle is expanded to include its generation number as well as the
%identifier for the underlying generational event.  In addition to the triggered-or-not state and list of
%dependent operations
%for the most recent generation, the owning node also remembers how many previous generations have already
%been triggered.  (There is no need to keep a list of dependent operations on previous generations.  Any
%new operation that is dependent on a generation that is known to be already triggered can immediately be
%executed.)  When an event is created, any generational event for whom the most recent generation has 
%triggered can be reused - the generation is increased by one and the state is returned to ``untriggered.''
%As before, this can be done with no inter-node communication.
Instead of attempting to free event data structures, our implementation aggressively recycles them.
Compared to a reference counting scheme our implementation requires fewer total event data structures
and has no client/runtime overhead.  The key observation is that one {\em generational event}
data structure can efficiently capture the state of one untriggered event and a very large 
number (e.g. $2^{32}-1$) of already-triggered events.  To accomplish this, we extend each event
handle to include its generation number as well as the identifier for its generational event.  Each
generational event remembers how many previous generations have already been triggered.  Any
new operation that is dependent on a generation that is known to already be triggered can immediately
be executed.  A generational event can be reused for a new generation as soon as the current generation has triggered. To
create a new event, a node finds a generational event in the triggered state,
increases the generation by one, and sets the generational event's state to untriggered.  As before, 
this can be done with no inter-node communication.

@@

Like events, deferred locks, or locks from here on, are created on demand by the node on which the creation request was made.  The lock
handles are also statically divided across the nodes, allowing the creation to be done without inter-node
communication.  Whereas events ownership is static, locks may migrate; the {\em creating}
node is the initial owner, but that ownership can be transferred to other nodes.

@@

When a lock request is made, an event is created to track when the grant occurs.  The current node then
examines its copy of the lock data structure to determine if it is the owner.
%(creating it if this the first operation for that lock on that node).  
If the current node is the owner and the lock isn't
held, the lock is granted immediately and the event is triggered.  If the lock is held, the event is added
to the list of local waiters.  Note that the event associated with
the lock request is the only data that must be stored.
%(This is the only thing that is stored.  The actual ``requestor'' of the lock
%is implicitly captured in which operations are dependent on the lock grant event.)  
If the current node isn't the owner, a {\em lock request} active message is
sent to the most recently known owner.  (If the receiver of a lock request message is no
longer the owner, it forwards the message on to the node it has recorded as the owner.)
%(If that information proves to be stale, the request is forwarded
%by the receiving node, eventually catching up to the current owner.)  
If the current owner's status shows
the lock is currently held, the bit for the requesting node is set in the remote waiters bitmask.  If the lock is
not held, the ownership of the lock is given to the requesting node via a {\em lock transfer} active
message.  A lock transfer message includes the bitmask of remote waiters and an up-to-date copy 
of the lock's payload.  The inclusion of the payload in the active message is the reason for the 
4KB size limit specified on payloads in Section~\ref{subsec:locks}.

@@

The first kind of reduction-only physical region is a {\em reduction fold instance}, which may only be
used for reduction operations that support a {\em fold} operation (see Listing~\ref{lst:regionapi}).  
A reduction fold instance is similar
to a normal physical region in that it is implemented as an array, indexed by the same element indices used
for the normal physical regions.  The difference is that each array element is a value of the reduction's
{\tt RHS} type which is generally the same size or smaller than the actual element type of the region 
(i.e. the {\tt LHS} type).  When the client requests a reduction operation to a particular location, 
the supplied right-hand-side value is folded in
to the corresponding location in the array.  An arbitrary number of reductions can be folded into each location.
When the reduction fold instance is copied back to a normal instance, the contents of the reduction instance
are transferred in a large block to the location of the normal instance.  On the receiving side
the runtime automatically applies the reduction instance to the destination region
by invoking the reduction function once for each location in the region via a cache-friendly 
linear sweep over the memory.
